---
title: "Beating the booky with xgBoost [1]"
author: "Tobias Diederich"
date: "18 März 2016"
output: html_document
---

Beating the booky in prediction performance is an ambitious goal. Odds calculating companies (further refered as bookys) put a huge effort in calculating accurate probabilities. I expect them to have large statistic branches with a lot of manual predicting support from sport analysts. Never the less, there are conditions, which could be exploited to gain an advantage: 
1. Market behaviour: Bookys adjust their odds depending on market behaviour, which is expected to be irrational  
2. Feature variety: Using different kinds of features than bookies  
3. Domain limitation: Concentrating on German 1. Bundesliga  
  
  
This blog is meant to be a continous story. In this first post I build a model with features, solely engineered from marketprices of participating players. In future work these features will be supplemented by shortterm form values for players, engineered from grades given by the famous german sportsmagazin 'Kicker'.  
As the modeling algorithm xgboost is used, because it is known to be high perfoming for different problems. Several winning solutions on kaggle used it.

## Goal

This work aims on implementing a model to predict accurate probabilities, which applied in a betting strategy against booky odds would simulate a preferably high percentage profit. This simulated percentage profit is implemented as a custom metric and integrated in the model tuning process. Caret provides a method to easily apply such custom metrics.



In my approach I use player features engineered for participating players in matches and aggregate them to match features for a categorial model with the outcome HOMEVICTORY, DRAW and VISITORSVICTORY. As player features, the players current marketprices and generated form values are used. Additional match features could contain Team formations. My cross validated predictions are compared with the predictions bookys made and comparing metrics are engineered. The most important metric I use in model tuning is the percentage profit I would have made, assumed I consequently applyed my predictions in betting against the parsed booky odds.

Feel free to check it out on github TODO.

## Data

The match, team and player data are collected from [transfermarkt.de](http://www.transfermarkt.de/) and [kicker.de](http://www.kicker.de/). Booky odds are collected from [sfstats.net](http://de.sfstats.net/).

Relevant data will be of germanies 1. Bundesliga for seasons 2005-2015. 

```{r echo=FALSE, results='hide',message=FALSE}
library(ggplot2)
library(vioplot)
library(corrplot)
library(polycor)
library(RMySQL)
library(caret)
library(tidyr)
library(properties)
library(e1071)
library(pROC)
library(gridExtra)
library(magrittr)
library(MASS)
library(gbm)
library(bnclassify)
library(C50)
library(kernlab)
library(xgboost)
library(testthat)
library(Hmisc)
library(plyr)
library(dplyr)
```

```{r warning=FALSE, message=FALSE, cache=TRUE}
source(file = '../dataProvider/loadData.R', echo = FALSE, encoding = 'UTF-8')
toMatchday <- 34
seasons <- c('2005-2006', '2006-2007', '2007-2008', '2008-2009', '2009-2010', 
             '2010-2011', '2011-2012', '2012-2013', '2013-2014', '2014-2015')
leagues <- c('BL1')
data <- loadTrainingData(toMatchday, seasons, leagues)
matches <- data$matches
odds <- data$odds
stats <- data$stats
```
For reproducability the data object is saved in data/BL1_2005-2015.Rds.

### Datasets

matches -> contains all matches with some informative properties  
odds -> contains booky odds and probabilies for all matches  
stats -> an observation contains information for one player in one match  

Here is a brief exploration of the raw data:
```{r warning=FALSE, message=FALSE}
describe(select(matches, goalsHome, goalsVisitors, matchResult))
describe(select(odds, HomeVictory, VisitorsVictory, Draw))
describe(select(stats, fitPrice, position, playerAssignment, formation))
```


## Feature Engineering

The features I extract are the marketprices of participating players aggregated by team (Home, Visitors), grouped position (TW, DEF, MID, OFF) and aggregation method (min, max, avg, sum). The grouping and aggregation creates price features on a match level. Players are integrated who are in the starting lineup. For the players on bench extra features are engineered.

```{r warning=FALSE, message=FALSE, cache=TRUE}
source(file = '../featureEngineering/positionFeatureExtraction.R', echo = FALSE, encoding = 'UTF-8')
assignedPositions <- c('tw', 'def', 'def', 'def', 'mid', 'mid', 'mid', 
                       'mid', 'off', 'off', 'off', 'off', 'off')
relNormalAssignments <- c('DURCHGESPIELT', 'AUSGEWECHSELT')
priceFuncts <- c('min', 'max', 'avg', 'sum')
benchPriceFuncts <- c('max', 'avg')
priceFeaturedMatches <- extractMatchResultFeatures(stats, matches, assignedPositions, relNormalAssignments,
                                                   priceFuncts = priceFuncts, benchPriceFuncts = benchPriceFuncts)
colnames(priceFeaturedMatches)
features <- priceFeaturedMatches[, grepl('Price', colnames(priceFeaturedMatches))]
```
So we have `r length(features)` features and `r nrow(features)` observations.

### Feature correlation

```{r warning=FALSE, message=FALSE}
# Position and Function correlation for price features of home teams
priceHomeData <- features[, grepl('Home', colnames(features))]
colnames(priceHomeData) <- gsub('_Price', '', gsub('_Home', '', colnames(priceHomeData)))
priceCorrData <- cor(priceHomeData)
corrplot(priceCorrData, order = 'AOE')
## => the aggregation functions have a very high correlation between each other
```

TODO

### Correlation between goalDiff and selected features

```{r}
theme1 <- trellis.par.get()
theme1$plot.symbol$col = rgb(.2, .2, .2, .4)
theme1$plot.symbol$pch = 16
theme1$plot.line$col = rgb(1, 0, 0, .7)
theme1$plot.line$lwd <- 2
trellis.par.set(theme1)

# Correlation between the goalDiff and different aggregation 
# Functions of the Price of offensive or midfield players
offPriceCols <- grepl('Price', colnames(features)) & 
  (grepl('off', colnames(features)) | grepl('mid', colnames(features))) &
  grepl('Home', colnames(features)) & !grepl('Bench', colnames(features))
offPriceData <- features[, offPriceCols]
colnames(offPriceData) <- gsub('off_', '', gsub('_Price', '', gsub('Home_', '', colnames(offPriceData))))

featurePlot(x = offPriceData, y = priceFeaturedMatches$goalDiff,
            plot = 'scatter', layout = c(2, 2),
            type = c('p', 'smooth'), span = .5)
## => off-price-home-features are slightly correlated with goalDiff
```

TODO

## Model fitting and tuning

The target of the model tuning process is to find a model which maximizes the simulated profit [%]. This is a custom metric and implemented in the function betMetricsSummary. This function is integrated in the caret resampling and tuning process (caret is so great!!).

The applied model is extreme gradient boosting, a popular algorithm which is know to have very good and stable predictive performance in various different cases. Additionally it has a lot of fine tuning potential and supports parallel processing.

### Configuration of the fitting process

I use 5-fold cross validation first. 10-fold would be probably better, but calculation would take much longer. A static seed is set to make the results reproduceable.

```{r warning=FALSE, message=FALSE}
source(file = '../models/models.R', 
       echo = FALSE, encoding = 'UTF-8')
modelInput <- selectModelInput(priceFeaturedMatches)
seed <- 16450
customCvContr <- trainControl(method = 'cv', number = 5, classProbs = TRUE, 
                              summaryFunction = betMetricsSummary)

resultFormula <- as.formula('matchResult ~ . -matchId -goalsHome -goalsVisitors -goalDiff')
```

### Extreme Gradiant Boosing

Tuning Parameters: 
- nrounds (# Boosting Iterations)
- max_depth (Max Tree Depth), 
- eta (Shrinkage)
- gamma (Minimum Loss Reduction)
- colsample_bytree (Subsample Ratio of Columns) 
- min_child_weight (Minimum Sum of Instance Weight)

```{r warning=FALSE, message=FALSE, cache=TRUE}
# Set tuning parameter sets #
extrBoostGrid <- expand.grid(nrounds = (1:5)*100,
                             max_depth = c(2, 3),
                             eta = c(.01, .025, .05),
                             gamma = c(0),
                             colsample_bytree = 1,
                             min_child_weight = 1)

set.seed(seed)
priceXGBoostModel <- train(form = resultFormula, data = modelInput, 
                           method = 'xgbTree', trControl = customCvContr, 
                           tuneGrid = extrBoostGrid, metric = 'GainPerc',
                           objective = 'multi:softprob', num_class = 3, 
                           allowParallel = TRUE)

priceXGBoostModel$results[as.integer(rownames(
  priceXGBoostModel$results)) == as.integer(rownames(priceXGBoostModel$bestTune)), ]

trellis.par.set(caretTheme())
plot(priceXGBoostModel)
vioplot(priceXGBoostModel$resample$GainPerc, names = 'Profit [%]', col = 'green')
title('Violin Plot of Profit Percentage in resamples')
summary(priceXGBoostModel$resample$GainPerc)

# Non-resampled training performance
testPred <- predict(priceXGBoostModel, modelInput)
confMatrix <- confusionMatrix(testPred, reference = modelInput$matchResult)
confMatrix$overall[1:2]
```



# Results and Forecast

The results are not overwhelming. The final model is very unstable concerning the profit (high SD). The minimum requirement for applying the model on real bets is a positive GainPerc over all folds and at least > 5%.
To achieve this, I think it is necessary to integrate new features. Imaginable features could describe team formations or distance of the visiting team to travel. A disadvantage of featured based on marketprices of players is, that the marketprices are created twice a year, so they describe the long term quality of a player. A way to integrate short term form predictors of players is to engineer features based on the kicker.de grade associated to players in past matches.